%% sections/abstract.tex
%% Limite AJ: 250 parole. Attuale: ~180 parole — espandere con risultati numerici reali.

We present \AAS\ (AstDyn-Adaptive Symplectic), a fourth-order
geometric integrator for asteroid orbit propagation, built on a
symplectic Yoshida Drift-Kick-Drift kernel and an adaptive
Sundman-type step-size law based on the spectral radius of the local
gravitational potential Hessian.
The step function
\[
  \Delta t = \eps \,/\, \sqrt{\rho\!\left(\nabla^2\Phi(\mathbf{q})\right)}
\]
yields a time step proportional to $r^{3/2}$ for Keplerian motion,
naturally concentrating computational effort at perihelion while
preserving the geometric structure of the Hamiltonian flow.
This scaling is consistent with prior adaptive symplectic literature;
the specific contribution here is its Hessian-spectral-radius
formulation, geometric curvature interpretation, and direct reuse of
the same Hessian for analytical STM propagation.
The Yoshida fourth-order composition scheme ensures long-term
quasi-conservation of energy through a well-defined shadow Hamiltonian.
Harmonic-mean symmetrization of the step function preserves
discrete time-reversibility.
Time-reversibility is treated as a distinct property from
symplecticity: strict symplecticity is proven for the fixed-step
DKD kernel, while the adaptive map is interpreted through the
Sundman transformed system (Section~\ref{sec:symplecticity_proof}).
The State Transition Matrix (STM) is propagated analytically via the
potential Hessian, eliminating the standard $12N_{\mathrm{steps}}$
overhead of numerical STM approaches.
We benchmark \AAS\ against \saba, \rkf, and
IAS15 \citep{ReinSpiegel2015} on three test asteroids:
(1)~Ceres ($e=0.0784$), (99942)~Apophis ($e=0.191$),
and (3200)~Phaethon ($e=0.890$).
\AAS\ achieves relative energy errors $\dEE < 10^{-10}$ over
one-year integrations, with best values in the
$1.96\times10^{-13}$--$2.77\times10^{-14}$ range for
Ceres and Apophis, and $3.49\times10^{-14}$ for Phaethon.
For Phaethon at equal accuracy
($|\Delta E/E|\approx7.00315\times10^{-5}$),
\AAS\ reaches the crossover at $\varepsilon=10^{-3}$ with
$7.0644\times10^4$ function evaluations versus
$2.922\times10^4$ for fixed-step \saba\
(a factor $2.42$ more evaluations).
The integrator is implemented in the open-source \astdyn\ library,
while the full paper package, benchmark scripts, and figure data used in this
article are available at \url{https://github.com/ITALOccult/AAS---Paper---MNRAS}.
